---
title: "Exploring Census PUMS Data"
author: "Lasya Harini Banka"
format: html
execute:
  warning: false
  message: false
---

## Introduction

This project uses data from the U.S. Census Public Use Microdata Sample (PUMS) API. The goal is to create functions that download, clean, combine, summarize and plot Census data. I will first collect the variable information needed to understand the API results. I will then create functions for requesting data from one or more years. The process is shown indetail below with explanations.

## Setup

The packages below are used for API requests, data cleaning and time conversion.

```{r}

library(tidyverse)
library(httr)
library(jsonlite)
library(lubridate)
library(hms)

census_key <- Sys.getenv("CENSUS_KEY")

if (census_key == "") {
  stop("Census API key was not found.")}
```

## Data Processing

### Parsing Variable Values

The Census API uses codes for many variables. I first created a function that downloads the variable information for one year and keeps only the variables needed for this project.

```{r}
pums_variables <- c(
  "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
  "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX")

get_variable_metadata <- function(year) {
  valid_years <- 2021:2024
  if (length(year) != 1 || !year %in% valid_years) {
    stop("Year must be 2021, 2022, 2023, or 2024.")}

  if (year %in% c(2021, 2022)) {
    state_variable <- "ST"
  } else {
    state_variable <- "STATE"}

  variables_to_keep <- c(
    pums_variables,
    state_variable,
    "REGION",
    "DIVISION")

  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json")

  response <- GET(url)
  stop_for_status(response)

  response_text <- content(
    response,
    as = "text",
    encoding = "UTF-8")

  response_list <- fromJSON(
    response_text,
    simplifyVector = FALSE)

  all_variables <- response_list$variables
  missing_variables <- setdiff(variables_to_keep, names(all_variables))

  if (length(missing_variables) > 0) {
    stop(
      "Some required variables were not found: ",
      paste(missing_variables, collapse = ", "))}

  all_variables[variables_to_keep]}
```

The information is downloaded for all four years and saved in an RDS file.

```{r}
#| label: save-variable-metadata

if (!file.exists("data/census_variable_values.rds")) {
  years <- 2021:2024

  census_variable_values <- setNames(
    lapply(years, get_variable_metadata),
    as.character(years)
  )

  dir.create("data", showWarnings = FALSE)

  saveRDS(
    census_variable_values,
    "data/census_variable_values.rds")}
```

```{r}
census_variable_values <- readRDS(
  "data/census_variable_values.rds")

names(census_variable_values)
```

### Extracting Value Labels

This helper extracts the codes and labels for one variable. These labels will be used when converting categorical, time and geographic variables.

```{r}
get_value_labels <- function(year, variable) {
  year_information <- census_variable_values[[as.character(year)]]
  variable_information <- year_information[[variable]]

  if (is.null(variable_information)) {
    stop("Variable information was not found.")}
  values <- variable_information$values

  if (is.null(values)) {
    return(NULL)}

  if (!is.null(values$item)) {
    values <- values$item}

  unlist(values, use.names = TRUE)}
```

Check using the SEX variable.

```{r}
get_value_labels(2024, "SEX")
```

### Parsing an API Response

The next helper turns the result from GET() into a tibble. The first row of the response contains the column names.

```{r}
response_to_tibble <- function(response) {
  stop_for_status(response)

  response_text <- content(
    response,
    as = "text",
    encoding = "UTF-8")

  api_data <- fromJSON(response_text)
  column_names <- as.character(api_data[1, ])
  data_rows <- api_data[-1, , drop = FALSE]

  output <- as_tibble(data_rows)
  names(output) <- column_names

  output}
```

This request is used to check the helper.

```{r}
test_url <- paste0(
  "https://api.census.gov/data/2024/acs/acs1/pums",
  "?get=AGEP,PWGTP,SEX",
  "&for=state:37")

test_response <- GET(
  test_url,
  query = list(key = census_key))
test_data <- response_to_tibble(test_response)

head(test_data)
```

### Converting Time Variables

JWAP and JWDP are stored as coded time ranges. This helper finds the midpoint of one time range.

```{r}
time_midpoint <- function(time_label) {
  if (is.na(time_label) || grepl("N/A", time_label, ignore.case = TRUE)) {
    return(NA_real_)}

  clean_label <- gsub("\\.", "", time_label)
  time_parts <- strsplit(clean_label, " to ", fixed = TRUE)[[1]]

  if (length(time_parts) != 2) {
    return(NA_real_)}

  start_time <- parse_date_time(
    trimws(time_parts[1]),
    orders = c("I:M p", "I p"))

  end_time <- parse_date_time(
    trimws(time_parts[2]),
    orders = c("I:M p", "I p"))

  if (is.na(start_time) || is.na(end_time)) {
    return(NA_real_)}

  start_seconds <- hour(start_time) * 3600 + minute(start_time) * 60
  end_seconds <- hour(end_time) * 3600 + minute(end_time) * 60

  if (end_seconds < start_seconds) {
    end_seconds <- end_seconds + 86400}

  midpoint <- (start_seconds + end_seconds) / 2
  midpoint %% 86400}
```

This function applies the midpoint calculation to a full column.

```{r}
convert_time_variable <- function(values, year, variable) {
  labels <- get_value_labels(year, variable)
  time_labels <- unname(labels[match(values, names(labels))])

  midpoint_seconds <- vapply(
    time_labels,
    time_midpoint,
    numeric(1))

  as_hms(midpoint_seconds)}
```

### Converting Categorical Variables

Categorical variables are changed from codes into factors with readable labels.

```{r}
convert_to_factor <- function(values, year, variable) {
  labels <- get_value_labels(year, variable)

  factor(
    values,
    levels = names(labels),
    labels = unname(labels))}
```

### Checking Geographic Values

The user can enter a geographic code, label or the word all.

```{r}
find_geography_code <- function(year, geography, geography_value) {
  if (tolower(geography_value) == "all") {
    return("*")}

  if (geography == "State") {
    if (year %in% c(2021, 2022)) {
      geography_variable <- "ST"
    } else {
      geography_variable <- "STATE"
    }
  } else {
    geography_variable <- toupper(geography)}

  labels <- get_value_labels(year, geography_variable)

  if (geography_value %in% names(labels)) {
    return(geography_value)}

  label_match <- which(
    tolower(unname(labels)) == tolower(geography_value))

  if (length(label_match) == 1) {
    return(names(labels)[label_match])}

  stop("The selected geographic value is not valid for ", geography, ".")}
```

### Requesting One Year of Data

The main function checks the inputs, creates the API request and cleens the returned columns.

```{r}
get_census_data <- function(
    year = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geography_value = NULL
) {
  if (length(year) != 1 || !year %in% 2021:2024) {
    stop("Year must be one value: 2021, 2022, 2023, or 2024.")}

  allowed_numeric <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")

  if (length(numeric_vars) == 0 ||
      !all(numeric_vars %in% allowed_numeric)) {
    stop("At least one valid numeric variable is required.")}

  if (!"PWGTP" %in% numeric_vars) {
    numeric_vars <- c(numeric_vars, "PWGTP")}

  if (length(setdiff(numeric_vars, "PWGTP")) < 1) {
    stop("Select at least one numeric variable besides PWGTP.")}

  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")

  if (length(categorical_vars) < 1 ||
      !all(categorical_vars %in% allowed_categorical)) {
    stop("Select at least one valid categorical variable.")}

  allowed_geographies <- c("State", "Division", "Region")

  if (length(geography) != 1 ||
      !geography %in% allowed_geographies) {
    stop("Geography must be State, Division, or Region.")}

  if (is.null(geography_value)) {
    if (geography == "State") geography_value <- "37"
    if (geography == "Division") geography_value <- "5"
    if (geography == "Region") geography_value <- "3"}

  if (length(geography_value) != 1) {
    stop("Only one geographic value can be selected.")}

  geography_code <- find_geography_code(
    year,
    geography,
    as.character(geography_value))

  requested_variables <- unique(c(numeric_vars, categorical_vars))

  base_url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums")

  response <- GET(
    base_url,
    query = list(
  get = paste(requested_variables, collapse = ","),
  `for` = paste0(tolower(geography), ":", geography_code),
  key = census_key))

  census_data <- response_to_tibble(response)

  regular_numeric <- setdiff(numeric_vars, c("JWAP", "JWDP"))

  for (variable in regular_numeric) {
    census_data[[variable]][census_data[[variable]] == "N"] <- NA
    census_data[[variable]] <- as.numeric(census_data[[variable]])}

  time_variables <- intersect(numeric_vars, c("JWAP", "JWDP"))

  for (variable in time_variables) {
    census_data[[variable]] <- convert_time_variable(
      census_data[[variable]],
      year, variable)}

  for (variable in categorical_vars) {
    census_data[[variable]] <- convert_to_factor(
      census_data[[variable]],
      year, variable)}

  geography_column <- names(census_data)[
    tolower(names(census_data)) == tolower(geography)]

  if (length(geography_column) == 1) {
    standard_name <- toupper(geography)
    names(census_data)[names(census_data) == geography_column] <- standard_name

    if (geography == "State") {
      if (year %in% c(2021, 2022)) {
        geography_variable <- "ST"
      } else {
        geography_variable <- "STATE"
      }
    } else {
      geography_variable <- standard_name}

    census_data[[standard_name]] <- convert_to_factor(
      census_data[[standard_name]],
      year,
      geography_variable
    )}

  class(census_data) <- c("census", class(census_data))
  census_data}
```

The default state is North Carolina which has Census code 37.

```{r}
north_carolina_data <- get_census_data(
  year = 2024,
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = "37"
)

head(north_carolina_data)
class(north_carolina_data)
```

### Combining Multiple Years

The final function calls the single-year function for each requested year, it adds a year column and combines the results.

```{r}
get_multiple_years <- function(
    years,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geography_value = NULL
) {
  if (length(years) < 1 || !all(years %in% 2021:2024)) {
    stop("Years must only contain 2021, 2022, 2023, or 2024.")}

  yearly_data <- vector("list", length(years))

  for (i in seq_along(years)) {
    current_year <- years[i]

    yearly_data[[i]] <- get_census_data(
      year = current_year,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geography_value = geography_value
    ) |>
      mutate(year = factor(current_year))}

  combined_data <- bind_rows(yearly_data)
  class(combined_data) <- unique(c("census", class(combined_data)))

  combined_data}
```

The example below shows combining of two years of North Carolina data.

```{r}
combined_nc_data <- get_multiple_years(
  years = c(2023, 2024),
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = "37"
)
head(combined_nc_data)
class(combined_nc_data)
```
### Testing More Function Options

The earlier examples used age and state code. I also checked that the function works with time variable and state name. I requested the usual arrival time at work. The JWAP values should be converted from Census codes into time values as below.

```{r}
time_test_data <- get_census_data(
  year = 2024,
  numeric_vars = c("JWAP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = "37"
)

time_test_data |>
  filter(!is.na(JWAP)) |>
  head()

class(time_test_data$JWAP)
```

The geography function also allows a geographic label instead of only numericcode. The example below uses the North Carolina label.

```{r}
state_label_test <- get_census_data(
  year = 2024,
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_value = "North Carolina/NC"
)

head(state_label_test)
```
## Summary and Plot Functions

Each row in the PUMS data may represent multiple people. The PWGTP column is therefore used as a weight when calculating summaries and creating plots.

### Census Summary Method

The function below calculates weighted means and standard deviations for numeric variables and weighted counts for categorical variables.

```{r}
summary.census <- function(
    object,
    numeric_vars = NULL,
    categorical_vars = NULL,
    ...
) {
  
  if (!"PWGTP" %in% names(object)) {
    stop("The data must contain PWGTP.")}
  
  if (!is.numeric(object$PWGTP)) {
    stop("PWGTP must be numeric.")}
  
  weights <- object$PWGTP
  
  if (is.null(numeric_vars)) {
    numeric_columns <- vapply(
      object,
      is.numeric,
      logical(1))
    
    numeric_vars <- setdiff(
      names(object)[numeric_columns],
      c("PWGTP", "JWAP", "JWDP"))}
  
  if (is.null(categorical_vars)) {
    categorical_columns <- vapply(
      object,
      is.factor,
      logical(1))
    
    categorical_vars <- names(object)[categorical_columns]}
  
  if (!all(numeric_vars %in% names(object))) {
    stop("One or more numeric variables were not found.")}
  
  if (!all(categorical_vars %in% names(object))) {
    stop("One or more categorical variables were not found.")}
  
  if (any(numeric_vars %in% c("PWGTP", "JWAP", "JWDP"))) {
    stop("PWGTP and time variables cannot be summarized as numeric variables.")}
  
  numeric_summary <- list()
  
  for (variable in numeric_vars) {
    
    if (!is.numeric(object[[variable]])) {
      stop(variable, " must be numeric.")}
    
    values <- object[[variable]]
    keep <- !is.na(values) & !is.na(weights)
    
    values <- values[keep]
    current_weights <- weights[keep]
    
    weighted_mean <- sum(
      values * current_weights
    ) / sum(current_weights)
    
    weighted_sd <- sqrt(
      sum(values^2 * current_weights) /
        sum(current_weights) -
        weighted_mean^2
    )
    
    numeric_summary[[variable]] <- list(
      mean = weighted_mean,
      sd = weighted_sd)}
  
  categorical_summary <- list()
  
  for (variable in categorical_vars) {
    
    if (!is.factor(object[[variable]]) &&
        !is.character(object[[variable]])) {
      stop(variable, " must be categorical.")}
    
    values <- droplevels(factor(object[[variable]]))
    keep <- !is.na(values) & !is.na(weights)
    
    categorical_summary[[variable]] <- tapply(
      weights[keep],
      values[keep],
      sum)}
  
  list(
    numeric = numeric_summary,
    categorical = categorical_summary)}
```

The function can summarize all available variables or only variables selected by the user.

```{r}
summary(north_carolina_data)

summary(
  north_carolina_data,
  numeric_vars = "AGEP",
  categorical_vars = "SEX")
```

### Census Plot Method

The plot method creates a weighted boxplot using one categorical variable and one numeric variable.

```{r}
plot.census <- function(
    x,
    categorical_var,
    numeric_var,
    ...
) {
  
  if (missing(categorical_var) || missing(numeric_var)) {
    stop("Select one categorical and one numeric variable.")}
  
  if (length(categorical_var) != 1 ||
      length(numeric_var) != 1) {
    stop("Select exactly one categorical and one numeric variable.")}
  
  if (!categorical_var %in% names(x)) {
    stop("The categorical variable was not found.")}
  
  if (!numeric_var %in% names(x)) {
    stop("The numeric variable was not found.")}
  
  if (!is.factor(x[[categorical_var]]) &&
      !is.character(x[[categorical_var]])) {
    stop("The categorical variable must be a factor or character variable.")}
  
  if (!is.numeric(x[[numeric_var]])) {
    stop("The numeric variable must be numeric.")}
  
  if (numeric_var %in% c("PWGTP", "JWAP", "JWDP")) {
    stop("PWGTP and time variables cannot be used as the numeric variable.")}
  
  ggplot(
    x,
    aes(
      x = .data[[categorical_var]],
      y = .data[[numeric_var]],
      weight = PWGTP
    )
  ) +
    geom_boxplot(fill = "lightblue") +
    labs(
      title = paste(numeric_var, "by", categorical_var),
      x = categorical_var,
      y = numeric_var
    ) +
    theme_minimal()}
```

The first plot compares age by sex. The second compares age across years.

```{r}
plot(
  north_carolina_data,
  categorical_var = "SEX",
  numeric_var = "AGEP")

plot(
  combined_nc_data,
  categorical_var = "year",
  numeric_var = "AGEP")
```

## Investigation

For this investigation, I looked at age and travel time to work in North Carolina from 2021 through 2024. I used AGEP for age and JWMNP for the number of minutes spent traveling to work. I also included sex and school enrollment as categorical variables.

### Getting the Investigation Data

The multiple-year function is used to collect all four years of data.

```{r}
investigation_data <- get_multiple_years(
  years = 2021:2024,
  numeric_vars = c("AGEP", "JWMNP", "PWGTP"),
  categorical_vars = c("SEX", "SCH"),
  geography = "State",
  geography_value = "37"
)

head(investigation_data)
dim(investigation_data)
```

Some people in the data did not travel to work. Their travel time is recorded as zero. I removed these observations before analyzing travel time.

```{r}
commute_data <- investigation_data |>
  filter(!is.na(JWMNP), JWMNP > 0)

class(commute_data) <- unique(
  c("census", class(commute_data))
)

head(commute_data)
dim(commute_data)
```

### Comparing the Summaries

I separated the first and last year to compare the weighted summaries for commuters in 2021 and 2024.

```{r}
investigation_2021 <- commute_data |>
  filter(year == "2021")

class(investigation_2021) <- unique(
  c("census", class(investigation_2021))
)

investigation_2024 <- commute_data |>
  filter(year == "2024")

class(investigation_2024) <- unique(
  c("census", class(investigation_2024))
)
```

The first summary uses both numeric variables and two categorical variables for 2021.

```{r}
summary_2021 <- summary(
  investigation_2021,
  numeric_vars = c("AGEP", "JWMNP"),
  categorical_vars = c("SEX", "SCH")
)

summary_2021
```

The same variables are summarized for 2024.

```{r}
summary_2024 <- summary(
  investigation_2024,
  numeric_vars = c("AGEP", "JWMNP"),
  categorical_vars = c("SEX", "SCH")
)

summary_2024
```

### Investigation Plots

The first plot compares the age distribution across all four survey years.

```{r}
plot(
  investigation_data,
  categorical_var = "year",
  numeric_var = "AGEP"
)
```

The second plot compares travel time to work for male and female commuters. Only positive travel times are used in this plot.

```{r}
plot(
  commute_data,
  categorical_var = "SEX",
  numeric_var = "JWMNP"
)
```
### Investigation Findings

The weighted average age of commuters was about 41.59 years in 2021 and 41.68 years in 2024. This shows that the average age changed very less between the two years. The weighted average travel time increased from about 24.5 minutes in 2021 to 25.7 minutes in 2024. This is an increase of about 1.18 minutes.
The age boxplots are similar across all four years. The medians and overall spread do not appear to change very much. The travel-time plot also shows similar distributions for male and female commuters, although both groups have several people with much longer travel times.

## Conclusion

In this project, I created functions to download and clean Census PUMS data from multiple years. I created a census class with summary and plot methods that use PWGTP as a weight.
The investigation compared age and travel time for North Carolina commuters. The average commuter age remained almost the same, while the average travel time was slightly higher in 2024 than in 2021. In conclusion we noted that the functions made it possible to request, combine summarize and visualize the Census data.